Higher order unification for free!

### 2024-04-08 paper/main.tex
more work on itro

% TODO: set this fields
%\setcopyright{cc}
%\setcctype{by}
\copyrightyear{2024}
\acmYear{XXXX 2024}
\acmBooktitle{YYY}
\acmDOI{ZZZZZZZZZZZZ}

@@

Specifying and implementing a logic or a proof system from scratch requires
significant effort.
Logical Frameworks and Higher Order Logic Programming Languages provide
dedicated, high-level Meta Languages (ML) to facilitate this task in two
key ways. The first and most well know one is that variable binding and
substitution can be taken for granted when ML binders represent object
logic ones. The second one that comes to mind is unification, the cornerstone
for proof construction and proof search, however in this paper we describe how
reusing that brick may not be as easy at is seems.

Meta languages such as Elf~\cite{elf}, Twelf~\cite{twelf},
$\lambda$Prolog~\cite{miller_nadathur_2012} and
Isabelle~\cite{10.1007/978-3-540-71067-7_7}
have been utilized to specify various
logics~\cite{felty88cade,books/sp/NipkowPW02,10.1007/BF00881873,felty93lics}.
In some cases, the most notable one being Higher Order Logic~\cite{books/sp/NipkowPW02},
the ML Isabelle is such a good fit that it implements an interactive proof
system for HOL, and not just a specification.

The object logic we are interested in is Coq's~\cite{Coq-refman}
Calculus of Inductive Constructions (CIC), and we want to implement a
type-class~\cite{wadler89} solver for Coq~\cite{Coq-refman} using
the Coq-Elpi~\cite{tassi:hal-01637063} meta programming language,
a dialect of $\lambda$Prolog already used to extend
Coq~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.
Type-class solvers are unification based proof search procedures
that  provide essential automation to widely used Coq libraries.
These solvers are reminiscent of Prolog: they back-chain lemmas taken
from a designated database of type class instances.

As an example we take the \coqIn{Decision} type class
from the Stdpp~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018}
library. The class
identifies predicates equipped with a decision procedure.
The following three designated \coqIn{Instances} state that:
1) the type \coqIn{fin n}, of natural numbers
smaller than \coqIn{n}, is \coqIn{Finite} (another type class);
2) the predicate \coqIn{nfact n nf}, linking a natural number
\coqIn{n} to its prime factors \coqIn{nf}, is decidable;
3) the universal closure of a predicate has a decision procedure if
its domain is finite and if the predicate is decidable.

\begin{coqcode}
Instance fin_fin n : Finite (fin n).              (* r1 *)
Instance nfact_dec n nf : Decision (nfact n nf).  (* r2 *)
Instance forall_dec A P : Finite A ~$\to$~             (* r3 *)
  ~$\forall$~x:A, Decision (P x) ~$\to$~ Decision (~$\forall$~x:A, P x).
\end{coqcode}

\noindent Given this database of instances a type-class solver is
able to prove the following statement automatically,
by back-chaining the lemmas above:

\begin{coqcode}
  Check _ : Decision (forall n: fin 7, nfact n 3).       ~\customlabel{goal:g}{(g)}~
\end{coqcode}

\noindent
The encoding of CIC provided by Elpi, that we will discuss at length later in
\cref{sec:encodings,sec:lang-spec}, is an Higher Order Abstract
Syntax (HOAS) datatype \elpiIn{tm} featuring (among others) the following
constructors:

@@

\noindent
Following standard $\lambda$Prolog~\cite{miller_nadathur_2012}
the concrete syntax to abstract, at the meta level, an expression
\elpiIn{e} over a variable \elpiIn{x}
is <<\elpiIn{x\ e}>>, and square brackets denote a list of
terms separated by comma. As an example we show the encoding of the Coq term
<<\coqIn{~$\forall$~y:t, nfact y 3}>>:

@@

\noindent
We now illustrate the encoding of the three instances above as higher-order
logic-programming rules: capital letters denote rule
parameters; \elpiIn{:-} separates the rule's head from the premises;
\elpiIn{pi w\ p} introduces a fresh nominal constant \elpiIn{w}
for the premise \elpiIn{p}.

\begin{elpicode}
finite (app[con"fin", N]).                            ~\customlabel{clause:r1}{(r1)}~
decision (app [con"nfact", N, NF]).                   ~\customlabel{clause:r2}{(r2)}~
decision (all A x\ app[P, x]) :- finite A,            ~\customlabel{clause:r3}{(r3)}~
  pi w\ decision (app[P, w]).
\end{elpicode}

\noindent
Unfortunately this translation of rule \ref{clause:r3} uses the
predicate \coqIn{P} as a first order term: for the meta
language its type is \elpiIn{tm}.
If we try to backchain the rule \ref{clause:r3} on the encoding of the goal
\ref{goal:g} given below

\begin{elpicode}
decision (all (app[con"fin", con"7"]) y\
  app[con"nfact", y, con"3"]).
\end{elpicode}

@@

\begin{elpicode}
app[con"nfact", y, con"3"] = app[P, y]                 ~\customlabel{problem:p}{(p)}~
\end{elpicode}

\noindent
In this paper we study a more sophisticated encoding of Coq terms allowing
us to rephrase the problematic rule \ref{clause:r3} as follows:
\todo{why not link?}

\begin{elpicode}
decision (all A x\ Pm x) :- decomp Pm P A, finite A,   ~\customlabel{clause:r3a}{(r3a)}~
  pi x\ decision (app[P, x]).
\end{elpicode}

@@

\begin{elpicode}
app[con"nfact", y, con"3"] = Pm y                     ~\customlabel{problem:pa}{(p')}~
Pm = x\ app[con"nfact", x, con"3"]     % assignment for Pm
A = app[con"fin", con"7"]              % assignment for A
\end{elpicode}

\noindent
After unifying the head of rule \ref{clause:r3a} with the goal, Elpi would run
the premise <<\elpiIn{decomp Pm A P}>> in charge of bringing the
assignment for \elpiIn{Pm} back to the domain \elpiIn{tm} of Coq terms:

\begin{elpicode}
P = lam A a\ app[con"nfact", a, con"3"]
\end{elpicode}

@@

\begin{elpicode}
decision (app[                    P                  , w])
decision (app[ lam A (a\ app[con"nfact", a, con"3"]) , w])
\end{elpicode}

@@

\begin{elpicode}
app[ lam A (a\ app[con"nfact", a, con"3"]) , x] =
app[ con"nfact"                            , N, NF]
\end{elpicode}
\noindent
~\\
The root cause of the problems we sketched in this example
is that the unification procedure \Ue of the meta language is not aware
of the equational theory of the object logic, even if both theories
include $\eta\beta$-conversion and admit most general
unifiers for unification problems in the pattern fragment \llambda~\cite{miller92jsc}.

\paragraph{Contributions}
In this paper we discuss alternative encodings of Coq in
Elpi (Section~\ref{sec:encodings}), then we identify a minimal language \Fo{}
in which the problems sketched here can be fully described.
We then detail an encoding \elpiIn{comp} from \Fo{} to \Ho (the language of
the meta language) and a decoding \elpiIn{decomp} to relate the unifiers
bla bla.. \noteE[inline]{citare Teyjus}.
The code discussed in the paper can be accessed at the URL:
\url{https://github.com/FissoreD/paper-ho}.

@@

The equational theory of Coq's Dependent Type Theory is very rich. In
addition to the usual $\eta\beta$-equivalence for functions, terms (hence types)
are compared up to proposition unfolding and fixpoint unrolling. Still,
for efficiency and predictability reasons, most form of automatic proof search
employ a unification procedure that captures a simpler one,
just $\eta\beta$, and that solves higher-order problems
restricted to the pattern fragment $\llambda$~\cite{miller92jsc}.
We call this unification procedure \Uo{}.

The equational theory of the meta language Elpi is strikingly similar,
since it it comprises $\eta\beta$ (for the meta language functions), and the
unification procedure \Ue{} solves higher-order problems in
$\llambda$.

In spite of the similarity the link between \Ue{} and \Uo{} is not trivial,
since the abstraction and application term constructors
the two unification procedures deal with are different. For example

\begin{tabular}{lcl}
\elpiIn{x\ f x} & \Ue{} & \elpiIn{f}\\
\elpiIn{lam A x\ app[con"f", x]} & \Uo{} & \elpiIn{con"f"}\\
\elpiIn{lam A x\ app[con"f", x]} & \nUe{} & \elpiIn{con"f"} \\
\elpiIn{P x} & \Ue{} & \elpiIn{x}\\
\elpiIn{app[P, x]} & \Uo{} & \elpiIn{x}\\
\elpiIn{app[P, x]} & \nUe{} & \elpiIn{x}\\
\end{tabular}

\noindent
One could ignore this similarity, and ``just'' describe the object language
unification procedure in the meta language, that is crafting a \elpiIn{unif}
predicate to be used as follows in rule \ref{clause:r3}:

\begin{elpicode}
decision X :- unif X (all A x\ app[P, x]), finite A,
  pi x\ decision (app[P, x]).
\end{elpicode}

\noindent
This choice would underuse the logic programming engine provided by
the meta-language since, by removing any datum from the head of rules,
indexing degenerates. Moreover the unification procedure \elpiIn{unif}
programmed in the meta-language is likely to be n order of magnitude
slower than one that is built-in.

To state precisely the problem we solve we need a \Fo{} representation
of CIC terms and a \Ho one.
We call \Eo the equality over ground terms in \Fo,
\Ee the equality over ground terms in \Ho,
\Uo the unification procedure we want to implement and
\Ue the one provided by the meta language.
\noteE[inline]{extend \Eo and \Ee with reflexivity on uvars.}

@@

\newcommand{\unifcomplete}[3]{
  % \forall #3_1 #3_2, %
    % \{#3_1, #3_2\} \subseteq \llambda \Rightarrow %
    #3_i \in \llambda \Rightarrow
      % \forall \rho, %
        \rho #3_1 #1 \rho #3_2  %
          \Rightarrow \exists \rho', #3_1 #2 #3_2 \mapsto \rho' \land \rho' \subseteq \rho
}

We write $t_1 \Ue t_2 \mapsto \sigma$ when $t_1$ and $t_2$ unify
with substitution $\sigma$; we write $\sigma t$ for the application of
the substitution to $t$, and $\sigma X = \{ \sigma t | t \in X\}$ when
$X$ is a set; we write $\sigma \subseteq \sigma'$ when $\sigma$ is more
general than $\sigma'$. We assume that the unification of our meta
language is correct:
%
\begin{gather}
  \unifcorrect{\Ee}{\Ue}{t} \label{prop:correct-ml}\\
  \unifcomplete{\Ee}{\Ue}{t}\label{prop:complete-ml}
\end{gather}

@@

We illustrate a compilation $\C{s}{t}{m}{l}$ that
maps a term $s$ in \Fo{} to a term $t$ in \Ho, a variable mapping $m$ and
list of links $l$.
The variable map connects unification variables in \Ho with variables
in \Fo{} and is used to ``decompile'' the assignment,
$\D{\sigma}{m}{l}{\rho}$. Links represent problematic sub-terms which
are linked to the unification variable that stands in their place in the
compiled term. These links are checked for or progress \noteE[inline]{XXX improve....}

We represent a logic program \emph{run} in \Fo{} as
a list \emph{steps} $p$ of length $\mathcal{N}$. Each step is a
unification problem between terms $\foUnifPb_{p_l}$ and
$\foUnifPb_{p_r}$ taken from the set of all terms \foUnifPb.
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_\mathcal{N}$.
\footnote{If the same rule is used multiple time in a run we
just consider as many copies as needed of the terms composing the
rules, with fresh unification variables each time}
The initial here $\rho_0$ is the empty substitution
%
\newcommand{\progress}{\ensuremath{\mathrm{progress}}\xspace}
\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}\xspace}
\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}\xspace}
\newcommand{\frun}{\ensuremath{\mathrm{frun}}\xspace}
\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}\xspace}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[5]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4 \land #5 = #3 \cup #4}}
\newcommand{\stepH}[6]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto (#5, #6)}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#3 \cup #4) \mapsto (#6',#5)}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3_{p-1}}{\sigma_{p}}{#3_p}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
%
$$
\begin{array}{l}
\stepF{\foUnifPb}{p}{\rho}{\rho''}
\deff
\stepFD{\foUnifPb}{p}{\rho}{\rho'}{\rho''}\vspace{2pt}\\
\runF{\foUnifPb}{\mathcal{N}}{\rho}
\deff
\runFD{\foUnifPb}{\mathcal{N}}
\end{array}
$$

We simulate each run in \Fo{} with a run in \Ho as follows.
Note that $\sigma_0$ is the empty substitution.
$$
\begin{array}{l}
\stepH{\hoUnifPb}{p}{\sigma}{\linkStore}{\sigma''}{\linkStore'} \deff\vspace{2pt}\\
  \qquad\stepHD{\hoUnifPb}{p}{\sigma}{\sigma'}{\sigma''}{\linkStore}\vspace{2pt}\\
\runH{\foUnifPb}{\mathcal{N}}{\rho} \deff \vspace{2pt}\\
  \qquad \hoUnifPb \times \mapStore \times \linkStore_0 = \{ (t_j,m_j,l_j) | s_j \in \foUnifPb, \C{s_j}{t_j}{m_j}{l_j} \}\vspace{2pt}\\
  \qquad \runHD{\hoUnifPb}{\mathcal{N}}{\linkStore}\vspace{2pt}\\
  \qquad \D{\sigma_{\mathcal{N}}}{\mapStore}{\linkStore_{\mathcal{N}}}{\rho_{\mathcal{N}}}
\end{array}
$$

\noindent
Here \hstep{} is made of two sub-steps: a call to \Ue (on the compiled
terms) and a call to \progress{} on the set of links. We claim the following:

\begin{proposition}[Simulation]\label{prop:simulation}
$\forall \foUnifPb, \forall \mathcal{N},$
$$
  \runF{\foUnifPb}{\mathcal{N}}{\rho}
  \Leftrightarrow
  \runH{\foUnifPb}{\mathcal{N}}{\rho}
$$
\end{proposition}

@@

\begin{proposition}[Simulation fidelity]\label{prop:fidelity}
In the context of~ \hrun, if $~\hoUnifPb \subseteq \llambda$ we have that
$\forall p \in 1 \ldots \mathcal{N},$
$$
\stepF{\foUnifPb}{p}{\rho_{p-1}}{\rho_p}
\Leftrightarrow
\stepH{\hoUnifPb}{p}{\sigma_{p-1}}{\linkStore}{\sigma_p}{\_}
$$
\end{proposition}
\noindent
In particular this property guarantees that a \emph{failure} in the \Fo{} run
is matched by a failure in \Ho{} \emph{at the same step}. We consider this
property very important from a practical point of view since it guarantees
that the execution traces are strongly related and in turn this enables a user
to debug a logic program in \Fo{} by looking at its execution trace in
\Ho{}.

XXX permuting hrun does not change the final result if check does not fail eagerly

XXX if we want to apply heuristics, we can apply them in decomp to avoid committing to
a non MGU too early

We can define $s_1 \Uo{} s_2$ by specializing the code of \hrun{} to
$\foUnifPb = \{ s_1, s_2 \}$ as follows:
%
$$
\begin{array}{l}
s_1 \Uo s_2 \mapsto \rho \deff \vspace{2pt}\\
\quad\C{s_1}{t_1}{m_1}{l_1} \land \C{s_2}{t_2}{m_2}{l_2}\vspace{2pt}\\
\quad    t_1 \Ue t_2 \mapsto \sigma' \land
    \progress(\{l_1,l_2\},\sigma') \mapsto (L,\sigma'') \land\vspace{2pt}\\
\quad \D{\sigma''}{\{m_1,m_2\}}{L}{\rho}
\end{array}
$$

>>>>

% TODO: set this fields
%\setcopyright{cc}
%\setcctype{by}
%\copyrightyear{2024}
%\acmYear{XXXX 2024}
%\acmBooktitle{YYY}
%\acmDOI{ZZZZZZZZZZZZ}

@@

Specifying and implementing a logic or a proof system from scratch requires
significant effort.
Logical Frameworks and Higher Order Logic Programming Languages provide
dedicated, high-level Meta Languages (ML) to facilitate this task in two
key ways. The first and most well know one is that variable binding and
substitution come for free is one uses the ML binders to represent the object
language ones. The second one that comes to mind is unification, a cornerstone
for proof construction and proof search. However reusing that brick may require
some additional work that we carry out in this paper.

Meta languages such as Elf~\cite{elf}, Twelf~\cite{twelf},
$\lambda$Prolog~\cite{miller_nadathur_2012} and
Isabelle~\cite{10.1007/978-3-540-71067-7_7}
have been utilized to specify various
logics~\cite{felty88cade,books/sp/NipkowPW02,10.1007/BF00881873,felty93lics}.
In the notable of Higher Order Logic~\cite{books/sp/NipkowPW02},
the ML Isabelle is such a good fit that it implements an interactive proof
system for the object logic, and not just a specification.
The object logic we are interested in is Coq's~\cite{Coq-refman}
Calculus of Inductive Constructions (CIC) and we want to implement a
form of proof search known as type-class~\cite{wadler89,sozeau08} resolution.
We want to use the Elpi~\cite{tassi:hal-01637063} meta programming language,
a dialect of $\lambda$Prolog already used to extend
Coq in various ways~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.
Type-class solvers are unification based proof search procedures
reminiscent of Prolog: they back-chain lemmas taken
from a designated database of ``type class instances''.
For this reason we believe that $\lambda$Prolog, hence Elpi, is a good fit for
implementing such as form of automation. In this paper we focus on one aspect of
this work, namely \emph{how to reuse the unification of the ML Elpi in order to
implement the one used by the type-class solver}.

We take as an example the \coqIn{Decision} and \coqIn{Finite} type classes
from the Stdpp~\cite{JUNG_KREBBERS_JOURDAN_BIZJAK_BIRKEDAL_DREYER_2018}
library. The class  \coqIn{Decision}
identifies predicates equipped with a decision procedure, while
\coqIn{Finite} the types whose inhabitants can be enumerated in a (finite) list.
The following three designated \coqIn{Instances} state that:
1) the type \coqIn{fin n}, of natural numbers
smaller than \coqIn{n}, is finite;
2) the predicate \coqIn{nfact n nf}, linking a natural number
\coqIn{n} to the number of its prime factors \coqIn{nf}, is decidable;
3) the universal closure of a predicate has a decision procedure if
its domain is finite and if the predicate is decidable.

\begin{coqcode}
Instance fin_fin: ~$\forall$~n, Finite (fin n).             (* r1 *)
Instance nfact_dec: ~$\forall$~n nf, Decision (nfact n nf). (* r2 *)
Instance forall_dec: ~$\forall$~A P, Finite A ~$\to$~            (* r3 *)
  ~$\forall$~x:A, Decision (P x) ~$\to$~ Decision (~$\forall$~x:A, P x).
\end{coqcode}

\noindent Given this database a type-class solver is expected to
prove the following statement automatically:

\begin{coqcode}
  Decision (~$\forall$~x: fin 7, nfact x 3)                   (* g *)
\end{coqcode}

\noindent
The proof found by the solver back-chains on rule 3 (the only rule
about the $\forall$ quantifier), and then solves the premises with rules
rules 1 and 2 respectively.
Note that rule 3 features a second order parameter \coqIn{P} that stands for
a function of type \coqIn{A ~$\to$~ Prop} (a predicate over \coqIn{A}).
The solver has to infer a value for \coqIn{P} by unifying the conclusion
of rule 3 with the goal, and in particular by solving the unification
problem \coqIn{P x = nfact x 3}. This higher order problem falls in the so
called pattern-fragment \llambda~\cite{miller92jsc} and hence admits a unique
solution \coqIn{P = ~$\lambda$~x.nfact x 3}.

In order to implement such a search in Elpi we shall describe the encoding
of CIC terms and then the encoding of rules. Elpi comes with
an Higher Order Abstract Syntax~\cite{10.1145/53990.54010} datatype of CIC
terms, called \elpiIn{tm}, that features (among others) the following
constructors:

@@

\noindent
Following $\lambda$Prolog~\cite{miller_nadathur_2012}'s standard syntax,
the meta level binding of a variable \elpiIn{x} in an expression
\elpiIn{e} is written <<\elpiIn{x\ e}>>, and square brackets denote a
list of terms separated by comma. For example the term
<<\coqIn{~$\forall$~y:t, nfact y 3}>> is encoded as follows:

@@

\noindent
We now illustrate the encoding of the three instances above as higher-order
logic-programming rules: capital letters denote rule
parameters; \elpiIn{:-} separates the rule's head from the premises and
\elpiIn{pi w\ p} introduces a fresh nominal constant \elpiIn{w}
for the premise \elpiIn{p}.

\begin{elpicode}
finite   (app [con"fin", N]).                         ~\customlabel{clause:r1}{(r1)}~
decision (app [con"nfact", N, NF]).                   ~\customlabel{clause:r2}{(r2)}~
decision (all A x\ app [P, x]) :- finite A,           ~\customlabel{clause:r3}{(r3)}~
  pi w\ decision (app [P, w]).
\end{elpicode}

\noindent
Unfortunately this intuitive encoding of rule \ref{clause:r3} does not work,
since it uses the predicate \coqIn{P} as a first order term: for the meta
language its type is \elpiIn{tm}. If we try to back-chain the rule
\ref{clause:r3} on the encoding of the goal g given below

\begin{elpicode}
decision (all (app [con"fin", con"7"]) y\              ~\customlabel{goal:g}{(g)}~
  app [con"nfact", y, con"3"]).
\end{elpicode}

@@

\begin{elpicode}
app [con"nfact", y, con"3"] = app [P, y]               ~\customlabel{problem:p}{(p)}~
\end{elpicode}

\noindent
In this paper we study a more sophisticated encoding of rules that, on a first
approximation, would shape \ref{clause:r3} as follows:

\begin{elpicode}
decision (all A x\ Pm x) :- link Pm P A, finite A,    ~\customlabel{clause:r3a}{(r3')}~
  pi x\ decision (app [P, x]).
\end{elpicode}

@@

\begin{elpicode}
app [con"nfact", y, con"3"] = Pm y                    ~\customlabel{problem:pa}{(p')}~
Pm = x\ app [con"nfact", x, con"3"]                   ~\customlabel{solution:pm}{(\rho)}~
\end{elpicode}

\noindent
Once the head of rule \ref{clause:r3a} unifies with the goal \ref{goal:g}
the premise <<\elpiIn{link Pm A P}>> brings the assignment \ref{solution:pm}
back to the domain \elpiIn{tm} of Coq terms:

\begin{elpicode}
P = lam A a\ app [con"nfact", a, con"3"]
\end{elpicode}

@@

\begin{elpicode}
decision (app[                    P                   , w])
decision (app[ lam A (a\ app [con"nfact", a, con"3"]) , w])
\end{elpicode}

@@

\begin{elpicode}
app[ lam A (a\ app [con"nfact", a, con"3"]) , x] =
app[ con"nfact"                             , N, NF]
\end{elpicode}
\noindent
~\\
The root cause of the problems we sketched in this example
is that the unification procedure \Ue of the meta language is not aware
of the equational theory of the object logic, even if both theories
include $\eta\beta$-conversion and admit most general
unifiers for unification problems in the pattern-fragment.

\paragraph{Contributions}
In this paper we identify a minimal language \Fo{} in which the problems
sketched in the introduction can be formally described.
We detail an encoding of a logic program on \Fo{} to a strongly related
logic program in \Ho (the language of the meta language) and we show that
the unification procedure of the meta language \Ue{} can be effectively
used to simulate a unification procedure \Uo for the object language that
features $\eta\beta$-conversion in the pattern-fragment.

\cref{sec:problem-statement} formally states the problem and gives the
intuition behind our solution. \cref{sec:encodings} discusses alternative
term encodings and related works. \cref{sec:lang-spec} introduces
the languages \Fo{} and \Ho{}, \cref{sec:simulation} describes a
basic simulation of higher order logic programs.
\cref{sec:eta,sec:invariant1} completes its equational theory
with support for $\eta$-conversion. \cref{sec:beta} deals with the
practical necessity of ``tolerating'' terms outside of the
pattern-fragment and discusses how heuristic can be applied.
Finally \cref{sec:implementation} discusses the implementation in Elpi.

The $\lambda$Prolog code discussed in the paper can be accessed at the
URL: \url{https://github.com/FissoreD/paper-ho}.

@@

@@

\newcommand{\unifcomplete}[3]{
  % \forall #3_1 #3_2, %
    % \{#3_1, #3_2\} \subseteq \llambda \Rightarrow %
    #3_i \in \llambda \Rightarrow
      % \forall \rho, %
        \rho #3_1 #1 \rho #3_2  %
          \Rightarrow \exists \rho', #3_1 #2 #3_2 \mapsto \rho' \land \rho' \subseteq \rho
}
\newcommand{\maybeeta}{\ensuremath{\Diamond\eta}\xspace}
\newcommand{\maybebeta}{\ensuremath{\Diamond\beta_0}\xspace}
\newcommand{\notllambda}{\ensuremath{\overline{\llambda}}\xspace}
%\newcommand{\notllambda}{\ensuremath{\Diamond\beta}\xspace}

The equational theory of Coq's Dependent Type Theory is very rich. In
addition to the usual $\eta\beta$-equivalence for functions, terms (hence types)
are compared up to proposition unfolding and fixpoint unrolling. Still,
for efficiency and predictability reasons, most form of automatic proof search
employ a unification procedure that captures a simpler one,
just $\eta\beta$, and that solves higher-order problems
restricted to the pattern fragment \llambda.
We call this unification procedure \Uo{}.

The equational theory of the meta language Elpi is strikingly similar,
since it it comprises $\eta\beta$ (for the meta language functions), and the
unification procedure \Ue{} solves problems in
\llambda as well.

In spite of the similarity the link between \Ue{} and \Uo{} is not trivial,
since the abstraction and application term constructors
the two unification procedures deal with are different. For example:
%
\setlength{\tabcolsep}{2pt}
\vspace{4pt}
\begin{tabular}{lcl|}
\elpiIn{x\ f x} & \Ue{} & \elpiIn{f}\\
\elpiIn{lam A x\ app[con"f", x]} & \Uo{} & \elpiIn{con"f"} \phantom{mn} \\
\elpiIn{lam A x\ app[con"f", x]} & \nUe{} & \elpiIn{con"f"}
\end{tabular}
\phantom{mn}
\begin{tabular}{lcl}
\elpiIn{P x} & \Ue{} & \elpiIn{x}\\
\elpiIn{app[P, x]} & \Uo{} & \elpiIn{x}\\
\elpiIn{app[P, x]} & \nUe{} & \elpiIn{x}\\
\end{tabular}
\vspace{4pt}

To state precisely the problem we solve we need a \Fo{} representation
of CIC terms and a \Ho one. We extend the equality over ground
terms to open ones with the reflexivity of unification variables (that is
unification variable X is equal to itself, not to another one such a Y).
We call \Eo that equality in \Fo{} and \Ee its counterpart in \Ho.
We call \Uo the unification procedure we want to implement and
\Ue the one provided by the meta language.
We assume that the unification of our meta
language is correct and complete in \llambda.
%
%\begin{gather}
%  \unifcorrect{\Ee}{\Ue}{t} \label{prop:correct-ml}\\
%  \unifcomplete{\Ee}{\Ue}{t}\label{prop:complete-ml}
%\end{gather}

We write $t_1 \Ue t_2 \mapsto \sigma$ when $t_1$ and $t_2$ unify
with substitution $\sigma$; we write $\sigma t$ for the application of
the substitution to $t$, and $\sigma X = \{ \sigma t ~|~ t \in X\}$ when
$X$ is a set; we write $\sigma \subseteq \sigma'$ when $\sigma$ is more
general than $\sigma'$ and we use set union to denote the concatenation of
two substitutions. We shall use $\rho$ for \Fo{} substitutions, and
$\sigma$ for the \Ho ones.

@@

We illustrate a compilation $\C{s}{t}{m}{l}$ that
maps a term $s$ in \Fo{} to a term $t$ in \Ho{} and produces
a variable mapping $m$ and list of links $l$.
The variable map connects unification variables in \Ho with variables
in \Fo{} and is used to ``decompile'' the assignment,
$\D{\sigma}{m}{l}{\rho}$.
Links are an accessory piece of information whose
description is deferred to \cref{sec:nutshell}.

We represent a logic program \emph{run} in \Fo{} as
a list \emph{steps} $p$ of length $\mathcal{N}$.
Each step $p$ represents 
a unification problem between two terms $\foUnifPb_{p_l}$ and
$\foUnifPb_{p_r}$ taken from the set of all terms \foUnifPb.
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_\mathcal{N}$.
\footnote{If the same rule is used multiple time in a run we
just consider as many copies as needed of the terms composing the
rules, with fresh unification variables each time}
%
\newcommand{\progress}{\ensuremath{\mathrm{progress}}\xspace}
\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}\xspace}
\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}\xspace}
\newcommand{\frun}{\ensuremath{\mathrm{frun}}\xspace}
\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}\xspace}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[5]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4 \land #5 = #3 \cup #4}}
\newcommand{\stepH}[6]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto (#5, #6)}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#3 \cup #4) \mapsto (#6',#5)}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3_{p-1}}{\sigma_{p}}{#3_p}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
%
$$
\begin{array}{l}
\stepF{\foUnifPb}{p}{\rho}{\rho''}
\deff
\stepFD{\foUnifPb}{p}{\rho}{\rho'}{\rho''}\vspace{2pt}\\
\runF{\foUnifPb}{\mathcal{N}}{\rho}
\deff
\runFD{\foUnifPb}{\mathcal{N}}
\end{array}
$$

\noindent We simulate each run in \Fo{} with a run in \Ho as follows.
%Note that $\sigma_0$ is the empty substitution.
$$
\begin{array}{l}
\stepH{\hoUnifPb}{p}{\sigma}{\linkStore}{\sigma''}{\linkStore'} \deff\vspace{2pt}\\
  \qquad\stepHD{\hoUnifPb}{p}{\sigma}{\sigma'}{\sigma''}{\linkStore}\vspace{2pt}\\
\runH{\foUnifPb}{\mathcal{N}}{\rho} \deff \vspace{2pt}\\
  \qquad \hoUnifPb \times \mapStore \times \linkStore_0 = \{ (t_j,m_j,l_j) | s_j \in \foUnifPb, \C{s_j}{t_j}{m_j}{l_j} \}\vspace{2pt}\\
  \qquad \runHD{\hoUnifPb}{\mathcal{N}}{\linkStore}\vspace{2pt}\\
  \qquad \D{\sigma_{\mathcal{N}}}{\mapStore}{\linkStore_{\mathcal{N}}}{\rho_{\mathcal{N}}}
\end{array}
$$

\noindent
Here \hstep{} is made of two sub-steps: a call to the meta language
unification and a check for \progress{} on the set of links.
We claim:

\begin{proposition}[Simulation]\label{prop:simulation}
$\forall \foUnifPb, \forall \mathcal{N},$ if $~\foUnifPb \subseteq \llambda$
$$
  \runF{\foUnifPb}{\mathcal{N}}{\rho}
  \Leftrightarrow
  \runH{\foUnifPb}{\mathcal{N}}{\rho}
$$
\end{proposition}

@@

\begin{proposition}[Simulation fidelity]\label{prop:fidelity}
In the context of$\;$ \hrun, if $~\foUnifPb \subseteq \llambda$ we have that
$\forall p \in 1 \ldots \mathcal{N},$
$$
\stepF{\foUnifPb}{p}{\rho_{p-1}}{\rho_p}
\Leftrightarrow
\stepH{\hoUnifPb}{p}{\sigma_{p-1}}{\linkStore_{p-1}}{\sigma_p}{\linkStore_p}
$$
\end{proposition}
\noindent
In particular this property guarantees that a \emph{failure} in the \Fo{} run
is matched by a failure in \Ho{} \emph{at the same step}. We consider this
property very important from a practical point of view since it guarantees
that the execution traces are strongly related and in turn this enables a user
to debug a logic program in \Fo{} by looking at its execution trace in
\Ho{}.

%\cref{sec:beta}
%XXX permuting hrun does not change the final result if check does not fail eagerly
% XXX if we want to apply heuristics, we can apply them in decomp to avoid committing to a non MGU too early

We can define $s_1 \Uo{} s_2$% by specializing the code of \hrun{} to
%$\foUnifPb = \{ s_1, s_2 \}$ 
as follows:
%
$$
\begin{array}{l}
s_1 \Uo s_2 \mapsto \rho \deff \vspace{2pt}\\
\quad\C{s_1}{t_1}{m_1}{l_1} \land \C{s_2}{t_2}{m_2}{l_2}\vspace{2pt}\\
\quad    t_1 \Ue t_2 \mapsto \sigma' \land
    \progress(\{l_1,l_2\},\sigma') \mapsto (L,\sigma'') \land\vspace{2pt}\\
\quad \D{\sigma''}{\{m_1,m_2\}}{L}{\rho}
\end{array}
$$

\begin{proposition}[Properties of \Uo{} in \llambda]
\begin{gather}
  \unifcorrect{\Eo}{\Uo}{s}\label{prop:correct}\\
  \unifcomplete{\Eo}{\Uo}{s}\label{prop:complete}
\end{gather}
\end{proposition}

\noindent
In addition to correctness and completeness in \llambda, we claim that \hrun
handles terms outside \llambda in the following sense:
